\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-30 syntactic-example.tex
reply to comments

We provide one example in its surface syntax as pseudocode, and their
meta-theory encoding. This example is computing the predecessor of a
mundane natural number, but we split the inductive definition into two
parts and making "O" and "S" standing alone.
"pred" is defined as $n \mapsto ("pred" n, n)$\YZ{looks like an ill-founded definition}, and thus a predecessor.

@@

 EndFamily. (* pN₁, handler₁, sig₃ *)
 pred@\YZ{Why is there no FRecursor?}@
  : self[A].N → self[A].N × self[A].N
  := ... 
  (* predT, pred₁, sig₄, sig₄', sl₄ *) 
 k := self[A].pred self[A].O 
(* pN₂, sig₅ *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.45\linewidth}
\begin{minted}[escapeinside=@@]{Coq}
Family A₂ extends A.
 Extend FInductive N := S : N → N.
 (* Nsig, sig₁₁, inh₁, O₂, sig₂₁, inh₂, 
    sig₂₂, sig₂₁', sig₂₂', sl₂₂, inh₃ *)
 Extend Family handler.
   Inherit O. (* inhᵢₙ₀, pN₃ *)
   S : self[A].N × self[A].N
       → self[A].N × self[A].N 
     := fun (x,y) => (y, self[A₂].S y).
 EndFamily. (* inhᵢₙ, sig₃₁, inh₄*)
 pred 
  : self[A₂].N → self[A₂].N×self[A₂].N 
  := ... 
  (* sig₄₁, inh₅, sl₄₁ *)
 Inherits k. 
 (* inh₆ *)
EndFamily.
    \end{minted}
  \end{minipage}
  \caption{Demonstrating Pseudo-Code}
\end{figure}

>>>>

We provide one example in its surface syntax as pseudocode, and their
meta-theory encoding. This example is computing the predecessor of a
mundane natural number, but we split the inductive definition into two
parts and making "O" and "S" standing alone.
"pred" will map $S~n \mapsto (n, S~n)$ and $O \mapsto (O, O)$, and thus a predecessor.

@@

 EndFamily. (* pN₁, handler₁, sig₃ *)
 FRecursor pred about self[A].N 
  motive (fun _ => self[A].N × self[A].N)
  using handler.
 (* pred
  : self[A].N → self[A].N × self[A].N *)
  (* predT, pred₁, sig₄, sig₄', sl₄ *) 
 k := self[A].pred self[A].O 
(* pN₂, sig₅ *)
EndFamily.
\end{minted}
  \end{minipage}
  \begin{minipage}[t]{0.45\linewidth}
\begin{minted}[escapeinside=@@]{Coq}
Family A₂ extends A.
 Extend FInductive N := S : N → N.
 (* Nsig, sig₁₁, inh₁, O₂, sig₂₁, inh₂, 
    sig₂₂, sig₂₁', sig₂₂', sl₂₂, inh₃ *)
 Extend Family handler.
   Inherit O. (* inhᵢₙ₀, pN₃ *)
   S : self[A].N × self[A].N
       → self[A].N × self[A].N 
     := fun (x,y) => (y, self[A₂].S y).
 EndFamily. (* inhᵢₙ, sig₃₁, inh₄*)
 Inherit pred.

  (* sig₄₁, inh₅, sl₄₁ *)
 Inherits k. 
 (* inh₆ *)
EndFamily.
    \end{minted}
  \end{minipage}
  \caption{Demonstrating Pseudo-Code}
\end{figure}

### 2022-06-30 syntax.tex
reply to comments

Of course we need to put an eye on the substitution law for each piece
of the syntax. Here $\bW \sigma$ is a large singleton
type~\cite{stone2000}, and it only has one ``element'', the inductive
type itself, inside. We formulate in this cumbersome way because we want
to have inductive type to be a field element and exposing the
constructors at the type level for later \textit{pattern-matching
exhaustiveness checking} (i.e. if an inductive type as field element has type
$\cU$, then no information about the inductive type is exposed).
\YZ{What is the distinction of 'W T' vs 'El (W T)'?}\EDJreply{(W T) is a term inside the type universe, El (W T) is really a type. They are basically the same thing but (W T) cannot be at the type position in a judgement, i.e. k |- q : (W T) is not allowed, we only have k |- q : El (W T). This is kind of standard notation (at least in both Sterling and Kaposi) so I will avoid explaining in the main text if you think ok?}\YZreply{I think explanation is warranted; readers unfamiliar with that line of work may find it confusing. BTW, is the elimination rule for "El (W T)" omitted?}\EDJreply{Please check the following paragraph for the explanation for El, and also Univ-1 is the elimination rule for "El T" in general. I just didn't write out the two equations c (El T) = T, El (c T) = T}\YZreply{Seems harmless to list these equations and move the reference to 'type universe a la Tarski' earlier.}\EDJreply{Done. Please check.}

@@

$$
\Rule[name=Ept Sig]
{}
{\goodSig{\Gamma}{\nu\cdot}{0}}
\quad
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}\YZ{How should I understand 'f' and 'A'?}\EDJreply{`f' is the seal, 'A' is the result of sealing. I think I wrote down the motivation of sealing... Is it still too hard? Do you think I need to think of a more friendly way?}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
$$

@@

$$
\Rule[name=Compile]
{}
{\goodType{\Gamma}{\cC \nu\cdot \equiv \top}{} 
\\ \goodTerm{\Gamma}{ \cCt \mu\cdot \equiv ()}{\cL \nu\cdot}
\\
\goodType{\Gamma}{\cC (\nu^+ \ \sigma \ \{f\} \ T) \equiv 
    \Sigma (\cC \sigma) T["sf" \ f]}{}
\\\\ \goodTerm{\Gamma}{\cCt \ (\mu^+ \ o \ \{f\}\ t) \equiv ((\cCt \ o), t["sf" \ f][(id, \cCt \ o)]) }{}
}
$$
% $$
% \Rule
% { \goodSeal{\Gamma}{f}{\sigma}{A}
% \\ \goodType{\Gamma, A}{T}{}
% \\ \goodTerm{\Gamma}{t}{T}
% }{}
% $$

$\goodSig{\Gamma}{\sigma}{n}$ means $\sigma$ is a well-typed signature
for linkages in the context $\Gamma$ with $n$ fields. We omit the
naming/label information for fields in the meta\-theory. We can use $\cL$
to retrieve the corresponding types of linkage of a given signature, and
$\cC$ to get the compiled linkage type (an existential type).\YZ{Why is Csigma needed? How is it different from Lsigma?}\EDJreply{It corresponds to the compilation of the family ``type'', which is part of the implementation detail. Basically we know every field of a family is dependent on the former fields. The implementation detail here is that, every field of a family is dependent on the compiled module of the former family (and thus includes all the former fields)} $\cL\sigma$ corresponds to the type of a family term, but $\cC\sigma$, accoding to \ruleref{Compile}, is really a deeply nested dependent pair. This is reflecting the implementation of our plugin : we know every field is defined in the context of the former fields, but our implementation actually make every field dependent on \textbf{the compiled module of the already-defined-part of the family}. Thus we need $\cC\sigma$ in our context.

The signature (for the linkage) and the linkage rules are quite similar to
the signature for inductive types.  They are both constructed from the
empty stuff, and respect substitution lemma for sure. We have
\ruleref{Sig Proj}, \ruleref{Lnkg Proj} projection for each component
due to harmony~\cite{pfenning2009lecture}, so we have local soundness
and local completeness ($\beta,\eta$ rules). We omit them and the
substitution laws for either field addition and projection. For
$\mu^+,\nu^+$ rules, we will omit $\{f\}$ if possible.\footnote{This
omission can be supported by Agda using Implicit Arguments}

@@

\begin{minted}[escapeinside=@@]{agda}
Seal : (Γ : Con) → (σ : Sig Γ n) → (A : Ty Γ) → Set 
Seal Γ A τ = Tm (Γ, 𝓒σ) A[π₁]@\YZ{tau?}\EDJreply{it should be pi. I fixed it now}@

@@

As we said earlier, $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is actually just
a shorthand for term judgement $\goodTerm{\Gamma , \cC
\sigma}{\_}{A[\pi_1]}$,\YZ{Not quite. Previously you used cL I think.}\EDJreply{Sorry. It is fixed. It should be $\cC$ in the context.}
a special form of \textit{local morphism}~\cite{abbott2003category}.
The substitution law should be understood as a lemma, and it is easily
derivable using $\gamma^\uparrow$.

The motivation for sealing judgement is acting like syntactic sugar
helpful for inheritance---due to the presence of inductive type in the
context, not everything is inheritable. For example, $ X_1 : \bW
\sigma_1 \vdash t : T$ in the parent family will generally not be able
to be inherited to the children once we override the inductive type $\bW
\sigma_1$ with $\bW \sigma_2$. In
other words, there is no way to ``transplant'' $t$ into the context $X_2
: \bW \sigma_2$.\YZ{Is sigma_2 the overriding of sigma_1? The reader has to guess.}\EDJreply{Thanks. Fixed}
This makes all the fields after this inductive type
non-inheritable---which shouldn't be the case because a lot of fields
might not be directly relying on this concrete $\bW \sigma_1$.

@@

$$
\Rule[name=Inh-Ext]
{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}
  \quad \goodTerm{\Gamma, A_2}{t}{T}\YZ{What is sigma2'?}\EDJreply{Sorry this should be A2, i.e. the result of sealing. Earlier version sigma2' is the result of sealing. It was type-set before the last meta-theory change}}
{\goodInh{\Gamma}{"inhext" \ h \ t}{\sigma_1}{(\nu^+ \  \sigma_2\  T)}}
\quad
\Rule[name=Inh-Inh]
{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}
\quad \goodType{\Gamma, A_1}{T}{}
\quad \goodTerm{\Gamma, A_2}{\uparrow^s}{A_1[\pi_1]}
}
{\goodInh{\Gamma}{"inhinh" \ h}{(\nu^+ \  \sigma_1 \  T)}{(\nu^+ \  \sigma_2 \  T[(\pi_1, \uparrow^s)])}}
$$
$$
\Rule[name=Inh+Inh]
{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}
\quad \goodTerm{\Gamma, A_2}{\uparrow^s}{A_1[\pi_1]}
\quad 
\goodInh{\Gamma, A_2}{i}{\tau_1[(\pi_1, \uparrow^s)]}{\tau_2}}
{\goodInh{\Gamma}{"inhnest" \ h \ i}{(\nu^+ \  \sigma_1\  \cL\tau_1)}{(\nu^+ \  \sigma_2\  \cL\tau_2)}}
\quad
$$

>>>>

Of course we need to put an eye on the substitution law for each piece
of the syntax. Here $\bW \sigma$ is a large singleton
type~\cite{stone2000}, and it only has one ``element'', the inductive
type itself, inside. We formulate in this cumbersome way because we want
to have inductive type to be a field element and exposing the
constructors at the type level for later \textit{pattern-matching
exhaustiveness checking} (i.e. if an inductive type as field element has type
$\cU$, then no information about the inductive type is exposed).

@@

$$
\Rule[name=Ept Sig]
{}
{\goodSig{\Gamma}{\nu\cdot}{0}}
\quad
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}\YZ{How should I understand 'f' and 'A'?}\EDJreply{`f' is the seal, 'A' is the result of sealing. I think I wrote down the motivation of sealing... Is it still too hard? Do you think I need to think of a more friendly way?}\YZreply{Yes, still too rough on the reader. I sense that this part is central to the metatheory, so it is worth explaining the details well.}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
$$

@@

$$
\Rule[name=Compile]
{}
{\goodType{\Gamma}{\cC \nu\cdot \equiv \top}{} 
\\ \goodTerm{\Gamma}{ \cCt \mu\cdot \equiv ()}{\cL \nu\cdot}\YZ{What is C_t? I don't think you talk about it anywhere.}
\\
\goodType{\Gamma}{\cC (\nu^+ \ \sigma \ \{f\} \ T) \equiv 
    \Sigma (\cC \sigma) T["sf" \ f]}{}
\\\\ \goodTerm{\Gamma}{\cCt \ (\mu^+ \ o \ \{f\}\ t) \equiv ((\cCt \ o), t["sf" \ f][(id, \cCt \ o)]) }{}
}
$$
% $$
% \Rule
% { \goodSeal{\Gamma}{f}{\sigma}{A}
% \\ \goodType{\Gamma, A}{T}{}
% \\ \goodTerm{\Gamma}{t}{T}
% }{}
% $$

$\goodSig{\Gamma}{\sigma}{n}$ means $\sigma$ is a well-typed signature
for linkages in the context $\Gamma$ with $n$ fields. We omit the
naming/label information for fields in the meta\-theory. We can use $\cL$
to retrieve the corresponding types of linkage of a given signature, and
$\cC$ to get the compiled linkage type (an existential type).
$\cL\sigma$ corresponds to the type of a family term, but $\cC\sigma$, accoding to \ruleref{Compile}, is really a deeply nested dependent pair. This is reflecting the implementation of our plugin : we know every field is defined in the context of the former fields, but our implementation actually make every field dependent on \textbf{the compiled module of the already-defined-part of the family}. Thus we need $\cC\sigma$ in our context.
\YZ{Why is Csigma needed? How is it different from Lsigma?}\EDJreply{It corresponds to the compilation of the family ``type'', which is part of the implementation detail. Basically we know every field of a family is dependent on the former fields. The implementation detail here is that, every field of a family is dependent on the compiled module of the former family (and thus includes all the former fields)}\YZreply{I don't think you should explain the metatheory from the angle of how things are implemented in your plugin. Rather, when describing your Coq plugin, you can say it matches the metatheory.
PS: you use two terminologies: 'compile' and 'seal', but I guess they mean very similar things. So avoid saying 'compile' here.
PPS: Sealing means hiding, so what is being hided? Why do you need to hide it? And how does the 'Compile' and 'Seal' rules perform hiding?}

The signature (for the linkage) and the linkage rules are quite similar to
the signature for inductive types.  They are both constructed from the
empty stuff, and respect substitution lemmas. We have
\ruleref{Sig Proj}, \ruleref{Lnkg Proj} projection for each component
due to harmony~\cite{pfenning2009lecture}, so we have local soundness
and local completeness ($\beta,\eta$ rules). We omit them and the
substitution laws for either field addition and projection. For
$\mu^+,\nu^+$ rules, we will omit $\{f\}$ if possible.\footnote{This
omission can be supported by Agda using Implicit Arguments}

@@

\begin{minted}[escapeinside=@@]{agda}
Seal : (Γ : Con) → (σ : Sig Γ n) → (A : Ty Γ) → Set 
Seal Γ A τ = Tm (Γ, 𝓒σ) A[π₁]@\YZ{tau?}\EDJreply{it should be pi. I fixed it now}\YZreply{Why is tau not used in the definition?}@

@@

As we said earlier, $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is actually just
a shorthand for term judgement $\goodTerm{\Gamma , \cC
\sigma}{\_}{A[\pi_1]}$,
a special form of \textit{local morphism}~\cite{abbott2003category}.
The substitution law should be understood as a lemma, and it is easily
derivable using $\gamma^\uparrow$.

The motivation for sealing judgement is acting like syntactic sugar
helpful for inheritance---due to the presence of inductive type in the
context, not everything is inheritable. For example, $ X_1 : \bW
\sigma_1 \vdash t : T$ in the parent family will generally not be able
to be inherited to the children once we override the inductive type $\bW
\sigma_1$ with $\bW \sigma_2$. In
other words, there is no way to ``transplant'' $t$ into the context $X_2
: \bW \sigma_2$.
This makes all the fields after this inductive type
non-inheritable---which shouldn't be the case because a lot of fields
might not be directly relying on this concrete $\bW \sigma_1$.

@@

$$
\Rule[name=Inh-Ext]
{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}
  \quad \goodTerm{\Gamma, A_2}{t}{T}}
{\goodInh{\Gamma}{"inhext" \ h \ t}{\sigma_1}{(\nu^+ \  \sigma_2\  T)}}
\quad
\Rule[name=Inh-Inh]
{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}
\quad \goodType{\Gamma, A_1}{T}{}
\quad \goodTerm{\Gamma, A_2}{\uparrow^s}{A_1[\pi_1]}
}
{\goodInh{\Gamma}{"inhinh" \ h}{(\nu^+ \  \sigma_1 \  T)}{(\nu^+ \  \sigma_2 \  T[(\pi_1, \uparrow^s)])}}
$$
$$
\Rule[name=Inh+Inh]
{\goodInh{\Gamma}{h}{\sigma_1}{\sigma_2}
\quad \goodTerm{\Gamma, A_2}{\uparrow^s}{A_1[\pi_1]}
\quad 
\goodInh{\Gamma, A_2}{i}{\tau_1[(\pi_1, \uparrow^s)]}{\tau_2}}
{\goodInh{\Gamma}{"inhnest" \ h \ i}{(\nu^+ \  \sigma_1\  \cL\tau_1)}{(\nu^+ \  \sigma_2\  \cL\tau_2)}}
\quad
$$
